Computing Vowel Elision in Yoruba Collocations

### 2023-04-14 paper.tex
Set date manually for latex
Since it thinks the date is January 1st 1980
Add initial parts of implementation for ot
Add recent changes, brush up deleting
Add introduction sec

\usepackage{caption}
\usepackage{graphicx}
\usepackage{tipa}
\usepackage{ot-tableau}
\usepackage[margin=1in]{geometry}

\usepackage{setspace}
% \doublespacing

@@

\title{Vowel Elision in Yoruba Collocations}
\author{Hemant Gouni}

@@

@@

\begin{figure}[h]
\caption{Deletion in Yoruba}
\label{fig:deletion-simple}
\begin{tabular}{c|c|c|c}
    \textbf{Underlying} & \textbf{Surface} & \textbf{Underlying} & \textbf{Surface}\\
    \textipa{/ow\'o-k\'i-ow\'o/} & \textipa{[ow\'ok\'ow\'o]} & \textipa{/\textdyoghlig{}\'o-\`Ew\`u/} & \textipa{[\textdyoghlig{}\'Ew\`u]}\\
    \textipa{/OmO-k\'i-OmO/} & \textipa{[OmOk\'OmO]} & \textipa{/ra-\`Og\`Ed\`E/} & \textipa{[r\`Og\`Ed\`E]}\\
    \textipa{/se-ol\'u/} & \textipa{[sol\'u]} & \textipa{/n\'i-oko/} & \textipa{[l\'oko]}\\
    \textipa{/\textdyoghlig{}E-ed\'e/} & \textipa{[\textdyoghlig{}ed\'e]} & \textipa{/s\'i-O\textdyoghlig{}\`a/} & \textipa{[s\'O\textdyoghlig{}\`a]}\\
\end{tabular}
\end{figure}

@@

The tableau in \pref{tableau:ssp-wins} shows that the language prefers to
delete a morpheme-initial vowel over a more sonorous vowel (\textipa{[\'e]}
rather than \textipa{[\'i]}). This ranking completely account for the pairs
shown in \pref{fig:deletion-except-1}. We would additionally like to rank
\onset{} and \ssp{} against each other, but with the data we have, they cannot
be. For valid surface forms, only deletion ever occurs in order to satisfy
\onset{} or \ssp. Every time a vowel is deleted, violations of either of these
constraints cannot increase (for \onset, because there's one less syllable
nucleus which requires an onset, and for \ssp, because there's one less
potentially non-low vowel). As a result, we cannot attempt a ranking between
\onset{} and \ssp{} either.

@@

\section{Computing Optimal Surface Forms}

@@

Our basic approach employs \textit{bounded depth-first search}. It is best to
illustrate this technique on a concrete example. Take, for instance, the
surface form \textipa{[sol\'u]} and its underlying representation
\textipa{/se-ol\'u/}. We will demonstrate our technique on this mapping.

>>>>

\usepackage{caption}
\usepackage{graphicx}
\usepackage{tipa}
\usepackage{ot-tableau}
\usepackage[margin=1in]{geometry}
\usepackage{minted}

\usepackage{setspace}
\onehalfspacing

@@

\title{Computing Vowel Elision in Yoruba Collocations}
\author{Hemant Gouni}
\date{May 15th, 2023}

@@

\section{Introduction}

The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
spoken primarily in West Africa, is well-documented. A large body of work
exists that explores the various aspects of it, largely from an
optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba}. Two
particularly well-studied instances of this problem are \textit{assimilation}
and \textit{deletion}. These are the two primary strategies employed by the
language for resolving a sequence of two or more vowels that occur at the
boundary of two adjoined morphemes. This text will specifically explore the
second strategy of deletion, and in doing so, will attempt to provide an
analysis that is simple enough to be feasible to verify computationally. That
is, once we have developed our analysis, the core contribution of this work
will be to encode it computationally so it can be quickly and easily verified
whether it works for a range of corpus data. We employ an optimality-theoretic
perspective throughout, and provide a thorough discussion in Section
\ref{sec:computing-optimal-surface-forms} of the prior work on computational
interpretations of optimality theory. Afterwards, we describe our own approach
to automating our analysis and discuss its efficacy.

@@

\begin{figure}[h]
\caption{Deletion in Yoruba}
\label{fig:deletion-simple}
\begin{tabular}{c|c|c|c}
    \textbf{Underlying} & \textbf{Surface} & \textbf{Underlying} & \textbf{Surface}\\
    \textipa{/ow\'o-k\'i-ow\'o/} & \textipa{[ow\'ok\'ow\'o]} & \textipa{/\textdyoghlig{}\'o-\`Ew\`u/} & \textipa{[\textdyoghlig{}\'Ew\`u]}\\
    \textipa{/OmO-k\'i-OmO/} & \textipa{[OmOk\'OmO]} & \textipa{/ra-\`Og\`Ed\`E/} & \textipa{[r\`Og\`Ed\`E]}\footnotemark\\
    \textipa{/se-ol\'u/} & \textipa{[sol\'u]} & \textipa{/n\'i-oko/} & \textipa{[l\'oko]}\\
    \textipa{/\textdyoghlig{}E-ed\'e/} & \textipa{[\textdyoghlig{}ed\'e]} & \textipa{/s\'i-O\textdyoghlig{}\`a/} & \textipa{[s\'O\textdyoghlig{}\`a]}\\
\end{tabular}
\end{figure}

\footnotetext{As you will see, this example contradicts our analysis. However,
it appears as a valid form only in a single source, so we note this and
continue as though it is an exception.}

@@

The tableau in \pref{tableau:ssp-wins} shows that the language prefers to
delete a morpheme-initial vowel over a more sonorous vowel (\textipa{[\'e]}
rather than \textipa{[\'i]}). This ranking completely accounts for the pairs
shown in \pref{fig:deletion-except-1}. We would additionally like to rank
\onset{} and \ssp{} against each other, but with the data we have, they cannot
be. For valid surface forms, only deletion ever occurs in order to satisfy
\onset{} or \ssp. Every time a vowel is deleted, violations of either of these
constraints cannot increase (for \onset, because there's one less syllable
nucleus which requires an onset, and for \ssp, because there's one less
potentially non-low vowel). As a result, we cannot attempt a ranking between
\onset{} and \ssp{} either.

@@

\label{sec:computing-optimal-surface-forms}
\section{Computing Optimal Surface Forms}

@@

Our basic approach employs \textit{bounded depth-first search}. It is best to
illustrate this technique on a concrete example. Take, for instance, the
surface form \textipa{[sol\'u]} and its underlying representation
\textipa{/se-ol\'u/} \textbf{continue with this running example}. We will
demonstrate our technique on this mapping.
